Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-08-27 paper.Rnw
edits (many more to add that are on a printed out version of the paper)

We apply LIME using six sets of tuning parameters to all observations in the \data \ test set to observe variability across tuning parameters. A quantile bin based simulation method is used for five of the LIME applications with the number of bins varying from 2 to 6 by application. We use 6 bins as the maximum, because the complexity of the explanations increases with the number of bins. The sixth application of LIME uses a kernel density simulation method. The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications. (See Appendix~\ref{lime-details} for descriptions of the tuning parameters.)

@@

plot_grid(sine_poi_exp, exp_scatter) 
@
\caption{(left) Visualization from the \emph{lime} R package of the LIME explanation for the \data \ prediction of interest. The bars represent the explainer model coefficients. (right) \emph{Explanation Scatterplot}. The points represent the simulated data colored by the random forest probabilities and sized by the assigned LIME weights. \hh{XXX could you make the complex prediction **much** lighter for this example? The diamond and the lines are hard to see, but should be the focus.} \kgc{Is that light enough? Feel free to edit the alpha value in the plot\_explain\_scatter function as you feel necessary. I also made the lines depicting the explainer model wider in limeaid, and there is an option 'line\_size' that can be adjusted.} The prediction of interest is shown as the diamond shaped point. The explainer model indicator variables associated with $x_1$ and $x_2$ (quantile bins containing the prediction of interest) are depicted by the solid lines. The line colors represent the explainer model coefficient signs and indicate whether the feature supports a prediction of `blue' (blue lines) or supports a prediction of `red' (red lines). This figure shows that the quantile bins are not flexible enough to capture the relationship between the random forest predictions and the $x_1$ and $x_2$ values.}
\label{fig:sine-exp-scatter}
\end{figure*}

@@

\paragraph{An Example LIME Explanation}

\kgc{I really like the idea of separating the description of the \emph{lime} figure from the explanation scatterplot, but the more I think about it, I don't like that it get it's own sub-section. I think it would be nicer to allocate the sub-sections to the visual diagnostic plots. Instead, I switched to using a heading like you created for introducing the sine data. I also changed the wording of the heading a bit. What do you think? Another option would be to drop the heading altogether and just say, "Here is an example of a LIME explanation...but it is not enough..."}

\hh{XXX I've just pulled the description of the left hand side of figure 4 out of the next section.} To demonstrate a LIME explanation, we focus on the misclassified point indicated in Figure \ref{fig:sine-plot-data} by the diamond as the prediction of interest. The left plot in Figure \ref{fig:sine-exp-scatter} shows a visualization of the explanation for the prediction of interest as suggested by Ribeiro et al and implemented in the \emph{lime} R package. The figure shows that the random forest returned a probability of \Sexpr{round(sine_poi_lime_default$label_prob[1], 2)} that the observation of interest belongs to category blue {(denoted as the ``label" in the figure).The lengths of the bars represent the explainer model coefficients associated with the indicator variables chosen by LIME as important. The color of the bar denotes the sign of the coefficient and whether the feature supports or contradicts a random forest classification of `blue'. The "explanation fit" value is the deviance ratio from the R package \emph{glmnet} \citep{simon:2011} for a ridge regression model. In other words, this is the $R^2$ value associated with the explainer model. In this case, it is \Sexpr{round(sine_poi_lime_default$model_r2[1], 2)} suggesting that the explainer model is not a good linear fit. However, it is commonly accepted that $R^2$ has limitations for assessing the quality of fit of a model \citep{sapra:2014}, and it should not be used as the only metric in a model assessment.

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-exp-scatter} can be interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]}, but the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]} provides support against a classification of blue. \hh{Because the support for blue by the value of feature $x_1$ outweighs the support for red by the value of feature $x_2$, the overall LIME explanation favors a label of blue for the prediction of interest.} %Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that the $x_1$ and $x_2$ locations of the prediction of interest provide more support for a random forest classification of blue over red.
This explanation agrees with the random forest probability of \Sexpr{sine_poi_lime_default$label_prob[1]} for blue. Note that both models classify the observation incorrectly. Based on the plot \hh{on the left of \autoref{fig:sine-exp-scatter}} alone, it is not possible to make an informed assessment of the explanation.

@@

\hh{For a further assessment of the explainer model}, we turn to an \emph{explanation scatterplot}: the \emph{explanation scatterplot} is a visual diagnostic for assessing the LIME claims of locality and fidelity for an individual explanation by juxtaposing the complex and explainer models in one plot. The format of an explanation scatterplot depends on the LIME simulation method. We introduce the explanation scatterplot here under the \emph{lime} R package default method in which the data are simulated uniformly from four quantile bins. In this scenario, LIME converts continuous predictor variables to indicator variables identifying whether the variable value falls in the same quantile bin as the prediction of interest or not. The indicator variables are used as the explainer model features.

>>>>

We apply LIME using six sets of tuning parameters to all observations in the \data \ test set to observe variability across tuning parameters. A quantile bin based simulation method \kge{(samples are simulated uniformly from a specified number of quantile bins)} is used for five of the LIME applications with the number of bins varying from 2 to 6 by application. We use 6 bins as the maximum, because the complexity of the explanations increases with the number of bins. \kge{Note that 4 quantile bins are the default method in \emph{lime}.} The sixth application of LIME uses a kernel density simulation method \kge{(samples are drawn from kernel density approximations of the feature distributions)}. The default methods for feature selection (forward selection) and the computation of the weights (Gower distance raised to an exponent of 1) are used for all applications. (See Appendix~\ref{lime-details} for descriptions of the tuning parameters.)

@@

plot_grid(sine_poi_exp, exp_scatter) 
@
\caption{(left) Visualization from the \emph{lime} R package of \kge{a} \sout{the} LIME explanation for the \data \ prediction of interest \kge{identified in \autoref{fig:sine-plot-data}}. \sout{The bars represent the explainer model coefficients.} (right) \emph{Explanation Scatterplot}. The points represent the simulated data colored by the random forest probabilities and sized by the assigned LIME weights. \hh{XXX could you make the complex prediction **much** lighter for this example? The diamond and the lines are hard to see, but should be the focus.} \kgc{Is that light enough? Feel free to edit the alpha value in the plot\_explain\_scatter function as you feel necessary. I also made the lines depicting the explainer model wider in limeaid, and there is an option 'line\_size' that can be adjusted.} The prediction of interest is shown as the diamond shaped point. The explainer model indicator variables associated with $x_1$ and $x_2$ (quantile bins containing the prediction of interest) are depicted by the solid lines. The line colors represent the explainer model coefficient signs and indicate whether the feature supports a prediction of `blue' (blue lines) or supports a prediction of `red' (red lines). This figure shows that the quantile bins are not flexible enough to capture the relationship between the random forest predictions and the $x_1$ and $x_2$ values.}
\label{fig:sine-exp-scatter}
\end{figure*}

@@

\paragraph{A Visual Representation of a LIME Explanation}

\kgc{I really like the idea of separating the description of the \emph{lime} figure from the explanation scatterplot, but the more I think about it, I don't like that it get it's own sub-section. I think it would be nicer to allocate the sub-sections to the visual diagnostic plots. Instead, I switched to using a heading like you created for introducing the sine data. I also changed the wording of the heading a bit. What do you think? Another option would be to drop the heading altogether and flow into the next paragraph.} \hh{XXX I've just pulled the description of the left hand side of figure 4 out of the next section.}

\kge{Before introducing the visual diagnostics, let us consider a commonly used visualization of a LIME explanation that is shown in Figure \ref{fig:sine-exp-scatter} (left). The plot is created using the \emph{lime} R package and provides a visual representation of a LIME explanation for the prediction of interest indicated in \autoref{fig:sine-plot-data} (when 4 quantile bins are used to simulate the data).} \sout{To demonstrate a LIME explanation, we focus on the misclassified point indicated in Figure \ref{fig:sine-plot-data} by the diamond as the prediction of interest. The left plot in Figure \ref{fig:sine-exp-scatter} shows a visualization of the explanation for the prediction of interest as suggested by Ribeiro et al and implemented in the \emph{lime} R package.}

The \kge{plot} \sout{figure} \kge{reports \Sexpr{round(sine_poi_lime_default$label_prob[1], 2)} as} \sout{shows that} the random forest \sout{returned a} probability \sout{of \Sexpr{round(sine_poi_lime_default$label_prob[1], 2)}} that the observation of interest belongs to category blue {(denoted as the ``label" in the \kge{plot \sout{figure}}). The lengths of the bars represent the \kge{\sout{explainer model}} coefficients \kge{from a ridge regression model (the explainer model) fit using the R package \emph{glmnet} \citep{simon:2011}} associated with the indicator variables chosen \kge{via feature selection \sout{by LIME as important}}. The color of \kge{a \sout{the}} bar denotes the sign of the coefficient and whether the feature "supports" or "contradicts" a random forest classification of `blue'. The "explanation fit" \kge{\sout{value}} is the deviance ratio from \kge{\sout{the R package}} \emph{glmnet} \kge{\sout{\citep{simon:2011} for a ridge regression model}}. In other words, this is the $R^2$ value associated with the explainer model. In this case, it is \Sexpr{round(sine_poi_lime_default$model_r2[1], 2)} suggesting that the explainer model is not a good linear fit. However, it is commonly accepted that $R^2$ has limitations for assessing the quality of fit of a model \citep{sapra:2014}\kge{\sout{,}} and \kge{\sout{it}} should not be used as the only metric in a model assessment.

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-exp-scatter} can be interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]}, but the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]} provides support against a classification of blue. \kge{Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that $x_2$ plays a more important role in the random forest prediction.} \kge{Additionally, consider that since the support for blue by $x_2$ outweights the support for red by $x_1$, the explainer model overall favors a label of blue for the prediction of interest, which} agrees with the random forest probability of \Sexpr{sine_poi_lime_default$label_prob[1]} for blue. Note that both models classify the observation incorrectly.

%\hh{Because the support for blue by the value of feature $x_1$ outweighs the support for red by the value of feature $x_2$, the overall LIME explanation favors a label of blue for the prediction of interest.} \kgc{I don't think this statement is quite right.}

%Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that the $x_1$ and $x_2$ locations of the prediction of interest provide more support for a random forest classification of blue over red.

@@

\kgc{I moved the next sentence here from the previous paragraph.} Based on the plot \hh{on the left of \autoref{fig:sine-exp-scatter}} alone, it is not possible to make an informed assessment of the explanation. \hh{For a further assessment of the explainer model}, we turn to an \emph{explanation scatterplot}: the \emph{explanation scatterplot} is a visual diagnostic for assessing the LIME claims of locality and fidelity for an individual explanation by juxtaposing the complex and explainer models in one plot. The format of an explanation scatterplot depends on the LIME simulation method. We introduce the explanation scatterplot here under the \emph{lime} R package default method in which the data are simulated uniformly from four quantile bins. In this scenario, LIME converts continuous predictor variables to indicator variables identifying whether the variable value falls in the same quantile bin as the prediction of interest or not. The indicator variables are used as the explainer model features.
